---
title: "PA1_template"
date: "September 29, 2017"
output:
  html_document: default
  pdf_document: default
---
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Exploratory Analysis of daily movement data from a personal activity monitoring device
 
 
## Introduction
 
It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the "quantified self" movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.
 
This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.
 
The data for this assignment can be downloaded from the course web site:   
- Dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) [52K]
 
The variables included in this dataset are:   
- steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)   
- date: The date on which the measurement was taken in YYYY-MM-DD format       
- interval: Identifier for the 5-minute interval in which measurement was taken
 
The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.
 
### Code for reading in the dataset and/or processing the data
```{r }
# set the working directory
#setwd("Desktop\\Coursera_Data_Science_Course\\Course 5 - Reproducible Research")
 
# Url of the data zip file
#url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
 
# Downlaod zip file
#download.file(url, destfile='activity.zip',method='curl')
 
# Unzip the data file. Use read.csv() to load the data set into R
activity <- read.csv("activity.csv")
summary(activity)
 
```
 
### What is mean total number of steps taken per day?
Note - For this part of the assignment, we will ignore the missing values in the dataset.
 
1. Calculate the total number of steps taken per day
```{r}
#Total num. of steps/day
ttl.steps.day <- aggregate(data=activity,steps~date,sum,na.rm = T)
 
head(ttl.steps.day)
```
 
2. Make a histogram of the total number of steps taken each day
```{r}
hist(ttl.steps.day$steps,
     xlab = "Total number of steps taken each day",
     main = "Histogram of the total number of steps taken each day",
     breaks = 10, col = "green")
abline(v = mean(ttl.steps.day$steps),col = "black",lwd = 2)
abline(v = median(ttl.steps.day$steps),col = "red",lwd = 2)
legend(x = "topright", c("Histogram", "Mean", "Median"),
       col = c("green", "black", "red"),
       lwd = c(2, 2, 2))
 
```
 
3. Calculate and report the mean and median of the total number of steps taken per day
```{r, }
mean(ttl.steps.day$steps)
median(ttl.steps.day$steps)
```
 
The values of mean and median are **`r as.integer(mean(ttl.steps.day$steps))`** and **`r median(ttl.steps.day$steps)`** respectively. The mean and median values are very close, as we can see in histogram also.
 
### What is the average daily activity pattern?
1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
 
```{r,warning=FALSE}
#calculating average steps/interval
avg.steps.interval <- aggregate(data=activity,steps~interval,mean,na.rm = T)
 
#Time series plot
library(ggplot2)
ggplot(avg.steps.interval,aes(c(interval),steps)) +
  geom_line(col = "steelblue") +
  labs(x = "Interval", y = "Average number of Steps") +
  ggtitle("Average number of Steps/Interval")
 
 
```
 
2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
 
```{r}
max.step.interval <- subset(avg.steps.interval,steps == max(steps))

max.step.interval
```
 
On average 5-minute interval number **`r max.step.interval$interval`** has the maximum number of steps **`r as.integer(max.step.interval$steps)`**.

###Imputing missing values

Note - There are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.  

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs).  
```{r}
#Totl number of record having NAs
sum(!complete.cases(activity))

```
Out of total **17,568** observations **`r sum(!complete.cases(activity))`** records have missing values.

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.   

We will replace the missing step values in dataset with the mean of num. of steps for that 5-minute interval.
```{r}

#We already have calculated the average steps/interval
avg.steps.interval <- aggregate(data=activity,steps~interval,mean,na.rm = T)

activity_no_na <- activity

#replacing NAs with mean 
for (i in 1:nrow(activity_no_na))
{
  if(is.na(activity_no_na$steps[i]))
  {
    #finding interval for missing step
    interval.to.search <- which(activity_no_na$interval[i] == avg.steps.interval$interval) 
    #assigning mean to missing step value
    activity_no_na$steps[i] <- avg.steps.interval[interval.to.search,]$steps
  }
}

```

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r,}
head(activity_no_na)

incomp.records <- sum(!complete.cases(activity_no_na))

```

In the new data set **activity_no_na** there are **`r incomp.records`** records with missing values.  

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?  
```{r}

#Average steps/day
avg.steps.day <- aggregate(data=activity_no_na,steps~date,sum)

#Histogram
hist(avg.steps.day$steps,
     xlab = "Total number of steps taken each day",
     main = "Histogram of the total number of steps taken each day",
     breaks = 10, col = "green")
abline(v = mean(avg.steps.day$steps),col = "black",lwd = 2)
abline(v = median(avg.steps.day$steps),col = "red",lwd = 2)
legend(x = "topright", c("Histogram", "Mean", "Median"),
       col = c("green", "black", "red"),
       lwd = c(2, 2, 2))

mean(avg.steps.day$steps)
median(avg.steps.day$steps)
```

The values of mean and median for new dataset are **`r as.integer(mean(avg.steps.day$steps))`** and **`r as.integer(median(avg.steps.day$steps))`** respectively. As we can observe that the mean and median values haven't changed much from before.  

### Are there differences in activity patterns between weekdays and weekends?  
1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r}
# converting date into weekdays
activity_no_na$weekday <- weekdays(as.Date(activity_no_na$date))

# classifying days names into weekday or weekend
for(i in 1:nrow(activity_no_na))
{
  if (activity_no_na$weekday[i] %in% c("Saturday", "Sunday"))
  {activity_no_na$weekday[i] = "Weekend"}
  else
  {activity_no_na$weekday[i] = "Weekday"}
}

#converting the column into a factor variable.
activity_no_na$weekday <- as.factor(activity_no_na$weekday)
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).  
```{r}
#Average steps/weekday + interval
avg.steps.weekday <- aggregate(data=activity_no_na,steps~interval+weekday,mean)

#Time series plot
library(ggplot2)
ggplot(avg.steps.weekday,aes(c(interval),steps)) +
  geom_line(col = "steelblue") + facet_grid(weekday ~ .) +
  labs(x = "Interval", y = "Number of Steps", title ="Average number of Steps/Interval") 
```

There seems to be no activity during 12:00 AM to 5:00 AM (0 to 500 on interval scale) as the subject might be sleeping during this time. From interval 500 (around 5:00 AM) the number of steps start increasing and peaked around 830 (around 8:30 AM) on both weekdays and Weekends, which suggest subject might be excersing or walk to his office.   
On weekdays on average the number of steps are low during 9:30 AM to 6:00 PM (930 to 1800 on interval scale) which suggests subject might be having a desk job.On Weekends the subject is more active than on weekdays as on average the number of steps are higher during 9:30 AM to 6:00 PM.  

